Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-05-14 Chapter-5/5-Preparatory.tex
s

We derive an estimate of the energy loss between the beamline momentum measurement and the TPC ($E_{loss}$) from the pion DDMC sample, since this quantity is not  measurable directly on data. 
The $E_{loss}$ distribution for the 60A  and 100A pion sample is shown in figure \ref{fig:ELoss60A}, left and right respectively. A clear double peaked structure is visible, which is due to the particles either missing or hitting the HALO paddle: a schematic rendering of this occurrence is  shown in figure \ref{fig:Halo}. The kinematic at WC4 determines the trajectory of a particle and whether or not it will hit the halo paddle. In figure \ref{fig:PxVsXTrue} , we plot the true  horizontal component of the momentum $P_x$ versus the true $X$ position at WC4 for pions missing the halo paddle (left) and for pions hitting the halo paddle (right) for the 60A MC simulation runs -- analogous plots are obtained with the 100A simulation. These distributions can be separated drawing a line in this position-momentum space. 
We use a logistic regression  \cite{agresti2013categorical}  as a classifier to find the best separating line, shown in both plots as the red line. We classify as ``hitting the halo paddle" all pions whose $P_x$ and $X$ are such that $$P_x +0.02* X - 0.4 < 0 $$ and as ``missing the halo  paddle" all pions whose $P_x$ and $X$ are such that $$P_x +0.02*X - 0.4 > 0, $$ where the coefficients of the line are empirically found by the logistic regression estimation. Overall, this simple methode classifies in the right category (hit or miss) about 86\% of the pion events.
We apply the same classifier on data. We assign  $E_{loss} = 32 \pm 4 $~MeV for pion events classified as ``hitting the halo paddle"; we assign  $E_{loss} = 24 \pm 3 $~MeV for pion events classified as ``missing the halo paddle".

@@

\section{Tracking Studies}\label{sec:TrackingStudies}
\subsection{Study of WC to TPC Match}\label{ch:WC2TPCMatchOptimization}

@@

\section{Tracking Studies}
In this section, we describe three studies. The first is a justification of the selection criteria for the beamline handshake with the TPC information.  We perform this study to boost  the correct identification of the particles in the TPC associated with the beamline information, while maintaining sufficient statistics for the cross section measurement. 
The second study is an optimization of the tracking algorithm, with the scope of maximizing the identification of the hadronic interaction point inside the TPC. These two studies are related, since the optimization of the tracking is performed on TPC tracks which have been matched to the wire chamber track; in turn, the tracking algorithm for TPC tracks determine the number of reconstructed tracks in each event used to try the matching with the wire chamber track. Starting with a sensible tracking reconstruction, we perform the WC2TPC matching optimization first, then the tracking optimization. The WC2TPC match purity and efficiency  are then calculated again with the optimized tracking.

@@

%In data, we attempt to uniquely match one WC-Track to one and only one reconstructed TPC track. This match is done by using in the $X$ and $Y$ coordinate of the extrapolated WC-Track to the upstream most point of the reconstructed TPC Track and by using the angle between the incoming track angle and the reconstructed TPC. We define $\Delta$X as the difference between the $x$ position of the most upstream point of the TPC track and the $x$ position of the WC track as projected to the TPC front face. $\Delta$Y is defined analogously. We define  $\Delta$R as $ \Delta \text{R} =  \sqrt{ \Delta \text{X}^2 +  \Delta \text{Y}^2}  $. The angle between the incident WC Track and the TPC track in the plane that contains them defines $\alpha$.

%We define a match between WC-track and TPC reconstructed track if  $\Delta \text{R} < r_{T}$, $\alpha < \alpha_{T}$ and the Z position of the first reconstructed point of the TPC track is within 2 cm from the TPC front face. The determination of the best $r_{T}$ and $\alpha_{T}$ is the scope of the following section.

%In MC, we mimic the matching between the WC and the TPC track on Monte Carlo by constructing a fake WC track using truth information at wire chamber four. We then apply the same WC to TPC matching algorithm as in data.

>>>>

We derive an estimate of the energy loss between the beamline momentum measurement and the TPC ($E_{loss}$) from the pion DDMC sample, since this quantity is not  measurable directly on data. 
The $E_{loss}$ distribution for the 60A  and 100A pion sample is shown in figure \ref{fig:ELoss60A}, left and right respectively. A clear double peaked structure is visible, which is due to the particles either missing or hitting the HALO paddle: a schematic rendering of this occurrence is  shown in figure \ref{fig:Halo}. The kinematic at WC4 determines the trajectory of a particle and whether or not it will hit the halo paddle. In figure \ref{fig:PxVsXTrue} , we plot the true  horizontal component of the momentum $P_x$ versus the true $X$ position at WC4 for pions missing the halo paddle (left) and for pions hitting the halo paddle (right) for the 60A MC simulation runs -- analogous plots are obtained with the 100A simulation. These distributions can be separated drawing a line in this position-momentum space. 
We use a logistic regression  \cite{agresti2013categorical}  as a classifier to find the best separating line, shown in both plots as the red line. We classify as ``hitting the halo paddle" all pions whose $P_x$ and $X$ are such that $$P_x +0.02* X - 0.4 < 0 $$ and as ``missing the halo  paddle" all pions whose $P_x$ and $X$ are such that $$P_x +0.02*X - 0.4 > 0, $$ where the coefficients of the line are empirically found by the logistic regression estimation. Overall, this simple methode classifies in the right category (hit or miss) about 86\% of the pion events. In MC, we assign  $E_{loss} = 32 \pm 4 $~MeV for pion events classified as ``hitting the halo paddle"; we assign  $E_{loss} = 24 \pm 3 $~MeV for pion events classified as ``missing the halo paddle". We apply the same classifier on data. A scan of the simulated geometry showed an excess of 3 cm of un-instrumented argon compared with the surveyed detector geometry. We account for this difference by assigning in data $E_{loss} = 24 \pm 6 $~MeV for pion events classified as ``hitting the halo paddle" and  $E_{loss} = 17 \pm 6 $~MeV for pion events classified as ``missing the halo paddle", where the uncertainty is derived as the standard deviation of the double peaked distribution.

The summary of the values for used for $E_{Loss}$ for the pion sample is listed in table \ref{tab:Eloss}  with the analogous results for the study on the kaon case.

\begin{table}[b]
\centering
\begin{tabular}{|l|c|c|}  
\hline
                          &  \multicolumn{2}{c|}{E$_{loss}$ [MeV]}    \\ \hline
                          & Hitting Halo          & Missing Halo     \\ \hline
Pion  MC           &  $32 \pm 4 $         &    $24 \pm 3$     \\ \hline
Pion Data          &  $25 \pm 6$          &    $17 \pm 6 $    \\ \hline
Kaon  MC          &  $37 \pm 5 $        &     $31 \pm 4 $    \\ \hline
Kaon Data         &  $26 \pm 6 $        &     $22 \pm 6 $    \\ \hline
\end{tabular}
\caption{Energy loss for pions and kaons.}
\label{tab:Eloss}
\end{table}

@@

\section{Tracking Studies}\label{sec:TrackingStudies}
In this section, we describe three studies. The first is a justification of the selection criteria for the beamline handshake with the TPC information.  We perform this study to boost  the correct identification of the particles in the TPC associated with the beamline information, while maintaining sufficient statistics for the cross section measurement.  The second study is an optimization of the tracking algorithm, with the scope of maximizing the identification of the hadronic interaction point inside the TPC.  These two studies are related, since the optimization of the tracking is performed on TPC tracks which have been matched to the wire chamber track; in turn, the tracking algorithm for TPC tracks determines the number of reconstructed tracks in each event used to try the matching with the wire chamber track. Starting with a sensible tracking reconstruction, we perform the WC2TPC matching optimization first, then the tracking optimization. The WC2TPC match purity and efficiency  are then calculated again with the optimized tracking.
The third study is an evaluation of  the angular resolution of the tracking algorithm in data and MC, which is particularly important in the context of the cross section analyses.

\subsection{Study of WC to TPC Match}\label{ch:WC2TPCMatchOptimization}
\subsection{Tracking Optimization}\label{ch:TrackOptimization}
\subsection{Study of WC to TPC Match}\label{ch:WC2TPCMatchOptimization}

\subsection{Angular Resolution}\label{sec:angleRes}
Scope of this study is to understand and compare the tracking performances and angular resolution of the TPC tracking on data and MC. 
We use the angular resolution of the tracking to determine  the value of smallest angle that we can reconstruct with a non-zero efficiency, effectively determining a selection on the angular distribution of the cross section measurement due to the tracking performance.

We start by selecting all the WC2TPC matched tracks used for the cross section analysis.  These tracks can contain from a minimum of 3 3D-space points to a maximum of 240  3D-space points.  We fit a line to all the 3D-space points associated with the track. 
For each track we calculate the average distance between each point in space and the fit line as follows 
\begin{equation} 
\bar d = \frac{\sum^N_i d_i}{N},
\end{equation} 
where $N$ is the number of 3D-space points of the track and $d_i$ is the distance of the $i$-th space point to the line fit. Several tests to compare the goodness of fit between data and MC have been considered. We decided to use $\bar d$ for its straightforward interpretation. The $\bar d$ distribution for data and MC is shown in Figure \ref{fig:Chi2AllPts} and shows a relatively good agreement between data and MC.

A visual representation of the procedure used to evaluate the angular resolution is shown in Figure \ref{fig:AngResProcedure}. 
For each track, we order the space points according to their Z position along the positive beam direction (panel a) and we split them in two sets: the first set contains all the points belonging to the first half of the track and the second set contains all the points belonging the second half of the track. We remove the last four points in the first set and the first four points in the second set, so to have a gap in the middle of the original track (panel b). We fit the first and the second set of points with two lines  (panel c). We then calculate the angle between the fit of the first and second half $\alpha$ (panel d). The angle $\alpha$ determines the spatial resolution of the tracking. The distributions for data and MC for $\alpha$ are given in Figure \ref{fig:trackingResolution}. The mean of the data and MC angular resolution are respectively

\begin{equation}
\bar\alpha_{Data} = (5.0 \pm 4.5) \text{ deg} 
\end{equation}

\begin{equation}
\bar\alpha_{MC} = (4.5 \pm 3.9) \text{ deg}. 
\end{equation}

Interaction angles smaller than the angle resolution are indistinguishable for the reconstruction. Therefore, we assess our ability to measure the cross section to be limited to interaction angles greater than 5.0 deg. More accurate studies of the angular resolution as a function of the kinetic energy and track length, albeit interesting, are left for an improvement of the analysis.

It is beneficial to take a moment to describe the definition of interaction angle. In case of elastic scattering, the definition is straightforward: the interaction angle is the angle between the incoming and outgoing pion, i.e.

\begin{equation}  
\theta = \cos^{-1} \Big(\frac{\vec p _{\text{incoming}}  \cdot\vec p _{\text{outgoing}}}{|\vec p _{\text{incoming}}|  |\vec p _{\text{outgoing}}| }\Big).
\end{equation}   
In case of inelastic scattering,  the presence of several topologies requires a more complex definition, as shown in figure \ref{fig:scatterPic}.  We define the scattering angle as the biggest of the angles between the incoming pion and the visible daughters, where the visible daughters are charged particles that travel more than 0.47 cm in the detector (see panel a); in case all the daughters are invisible, the angle is assigned to be 90 deg (see panel b). We chose this working definition of scattering angle for inelastic scattering keeping in mind how our tracking reconstruction works: the tracking will stop correctly in case of all the daughters are not visible in the detector and it is likely to stop correctly if multiple daughters form an interaction vertex. The only ``dangerous" case is the production of one charged daughter plus neutrals, which we can study with this working definition of scattering angle (see panel c).

We can see the effects of the angular resolution on the cross section by plotting the true Geant4 cross section for interaction angles greater than a minimum interaction angle. Figure \ref{fig:trueWithAngles} shows the  true Geant4 cross section  for interaction angles greater than 0 deg (green), 4.5 deg (red), 5.0 deg  (blue) and 9.0 deg (yellow). 
A small $0.5 \text{ deg}$ systematic shift between the mean of the data and MC angular resolution is present.%, which we account for in the context of the MC efficiency correction to the cross section, as presented in \ref{sec:angSys}.

\begin{figure}[ht]
\begin{minipage}[t]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{Chapter-5/Images/cDAvg.png}
\caption[]{Distributions of the average distance between each 3D point in space and the fit line,  $\bar d$ for the data used in the pion cross section analysis and the pion only DDMC. The distributions are area normalized. } \label{fig:Chi2AllPts}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{Chapter-5/Images/cTrackingDeg.png}
\caption[]{Distributions of angular resolution $\alpha$ for data used in the pion cross section analysis and pion only DDMC. The distributions are area normalized. } \label{fig:trackingResolution}
\end{minipage}
\end{figure}

\begin{figure}[ht]
\begin{minipage}[t]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{Chapter-5/Images/TrackingProcedure.png}
\caption{A visual representation of the procedure used to evaluate the angular resolution.}
\label{fig:AngResProcedure}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{Chapter-5/Images/Daughters.png}
\caption{A visual representation of the scattering angle definition in case of inelastic scattering.}
\label{fig:scatterPic}
\end{minipage}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=0.48\textwidth]{Chapter-5/Images/cTrueXSAngle.png}
\caption{ True ($\pi^-{-}, Ar$) cross section for interaction angles greater than 0 deg (green), 4.5 deg (red), 5.0 deg  (blue) and 9.0 deg (yellow). }
\label{fig:trueWithAngles}
\end{figure}

@@

\section{Tracking Studies}

@@
